\documentclass[12pt,a4paper]{article} % uncomment, if revtex is not installed
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{color}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{wasysym}
\usepackage{natbib}
\usepackage{fixmath}
\usepackage{comment}
\usepackage{fancyhdr}
\usepackage{pgfplots}
\usepackage{listings}
\pagestyle{fancyplain}
\usepackage{graphicx}
\usepackage{float}

\graphicspath{ {./} }
\fancyhf{}
\setlength{\parindent}{0in}
\lstset{
tabsize = 4, %% set tab space width
showstringspaces = false, %% prevent space marking in strings, string is defined as the text that is generally printed directly to the console
numbers = left, %% display line numbers on the left
commentstyle = \color[rgb]{0,0.5,0}, %% set comment color
keywordstyle = \color{blue}, %% set keyword color
stringstyle = \color{red}, %% set string color
rulecolor = \color{black}, %% set frame color to avoid being affected by text color
basicstyle = \small \ttfamily , %% set listing font and size
breaklines = true, %% enable line breaking
numberstyle = \tiny,
}

\lhead{ \fancyplain{}{Ivin Lee, il315@cam.ac.uk} }
\rhead{ \fancyplain{}{Databases}}
\cfoot{ \fancyplain{}{\thepage{}} }

\newcommand{\snt}[1]{\textcolor{magenta}{#1}} % edited by Sergei 
\newcommand{\crsid}[1]{\textcolor{red}{#1}} % edited by crsid
\newcommand{\sntdel}[1]{\textcolor{green}{\sout{#1}}} % deleted by Sergei

\newcommand{\problem}[1]{\subsection*{#1}
\setcounter{equation}{0}}
\newcommand{\question}[1]{\problem{Q. #1}}
\newcommand{\pproblem}[1]{\subsubsection*{(#1)}}

\author{Ivin Lee}
\title{Databases}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
\section{Introduction}
\subsection{Database Systems in course}
	\begin{itemize}
	\item HyperSQL
	\item DOCtor Who
	\item Neo4j
	\end{itemize}
\textbf{Interface}

This course will deal with Database Management Systems from the perspective of an application developer, and will teach the interfaces of these systems.
\\\\
\subsection{Database Management System}
Implements CRUD operations
	\begin{itemize}
	\item Create
	\item Read
	\item Update
	\item Delete
	\end{itemize}
Implements ACID transactions for concurrent updates
	\begin{itemize}
	\item Atomicity: either all actions carried out or none carried out
	\item Consistency: database is consistent before and after transactions
	\item Isolation: every transaction happens independently of other transactions: allows for concurrency
	\item Durability: if transaction happens successfully, then its effects will persist
	\end{itemize}

\subsection{SQL}
HyperSql - just a java file, overhead of using is low
\\\\
\subsection{NoSQL}
Not only SQL - non-relational, distributed, open-source, horizontally scalable.
	\begin{itemize}
	\item Scalability: data stored on multiple machines
	\item Fault Tolerance: service can survive failure of some machines
	\item Lower Latency: data located closer to widely distributed users
	\end{itemize}
\subsection{Distributing Data}
	\begin{itemize}
	\item Replication
	\item Partitioning
	\end{itemize}
\subsection{CAP Concepts}
	\begin{itemize}
	\item Consistency: all reads return data up-to-date
	\item Availability: all clients can fine some replica of the data
	\item Partition Tolerance: system can still operate despite loss or failure of part of the system
	\end{itemize}
Cannot achieve all of these, some balance.
\\\\
\subsection{CAP Principle}

In a highly distributed system, 
	\begin{itemize}
	\item Assume that network partitions and other connectivity problems will occur
	\item Implementing ACID transactions is very difficult and slow
	\item trade-off between availability and consistency
	\end{itemize}
Eventual consistency - if update activity ceases, then the system will reach a consistent state.
\\\\
\subsection{BASE}
	\begin{itemize}
	\item BA: basically available
	\item S: soft state
	\item E: eventual consistency
	\end{itemize}
This is an area of ongoing research.
\\\\
\subsection{Polyglot Persistence}

Using various Database Management Systems to manage different sets of data
\\\\
Conclusion: there will be a lot of changes in Database Management Systems.
\section{Entity-Relationship Diagrams}
\begin{itemize}
\item Entities (squares): nouns of the model
	\begin{itemize}
	\item Weak entities: 
		\begin{itemize}
		\item existence depends on existence of another entity
		\end{itemize}
	\item Sub-entities: inherit all attributes of the parent entity
	\end{itemize}
\item Attributes(ovals): represent properties
\item Key (underlined): attribute whose value uniquely identifies an entity instance
\item Relationship (diamond): verbs, relation between entities
	\begin{itemize}
	\item Cardinality:
		\begin{itemize}
		\item one-to-many: arrow to one
		\item many-to-one
		\item many-to-many
		\item one-to-one
		\end{itemize}
	\item May also have attributes
	\item Can be more than 2 parties, i.e. ternary relationship
	\end{itemize}
\end{itemize}
\section{Relational Algebra}
This will probably clash with the lectures on discrete mathematics but I'll still type them here
\subsection{Mathematical Relations}
Suppose S and T are sets. The Cartesian product is the set
\begin{align*}
S\times T=\{(s,t)|s\in S,t\in T\}
\end{align*}
A binary relation over $S\times T$ is any set $R$ with
\begin{align*}
&R\subseteq S\times T\\
\iff&(s,t)\in R\\
\iff& sRt
\end{align*}
S and T are domains. We are interested in finite relations R that can be stored.
\\\\
n-ary relations
\begin{align*}
R\subseteq S_1\times S_2\times...\times S_n=\{(s_1,s_2,...s_n)|s_i\in S_i\}
\end{align*}
Each relation is a row in a table. We can pair the column name (attribute name) with each element of $S_i$, so we can tell that a value $6$ in an element of a relation refers to, say, \verb|age: 6| or \verb|height: 6|. This way, column order doesn't matter.
\\\\
A database relation R is a finite set
\begin{align*}
R\subseteq \{\{(A_1,s_1),(A_2,s_2),...,(A_n,s_n)\}|s_i\in S_i\}
\end{align*}
We specify $R$'s schema as $R(A_1:S_1, A_2:S_2,...,A_n:S_n)$. Each element of $R$ is a \textbf{record}. Each $S_n$ could be a \verb|string|, \verb|integer| etc.
\subsection{Database Query Language}
Input: a collection of relation instances

Output: a single relation instance
\begin{align*}
R_1,R_2,...,R_k\Rightarrow Q(R_1,R_2,...,R_k)
\end{align*}
\subsection{Relational Algebra}
\begin{table}[H]
\begin{tabular}{lcrr}
Q & ::= & R             & base relation \\
  & |   & $\sigma_p(Q)$ & selection     \\
  & |   & $\pi_X(Q)$    & projection    \\
  & |   & $Q\times Q$   & product       \\
  & |   & $Q-Q$         & difference    \\
  & |   & $Q\cup Q$     & union         \\
  & |   & $Q\cap Q$     & intersection  \\
  & |   & $\rho_M(Q)$   & renaming     
\end{tabular}
\end{table}
$X=\{A_1, A_2, ...,A_k\}$ is a set of attributes.

$M=\{A_1\mapsto B_1,A_2\mapsto B_2,...,A_k\mapsto B_k\}$ is a renaming map
\\\\
A query must be well-formed: all column names must be distinct
\subsubsection{Selection}
Selecting one row:
\\\\
$\sigma_{A>12}(R)$

\verb|select distinct * from R where R.A > 12|
\\\\
Reason for \verb|distinct| is because SQL works with multisets so to follow set properties we need to use it. The choice of multisets is because there may be times when we need to sum these values, so we need to preserve duplicates.
\subsubsection{Projection}
Selecting multiple columns

$\pi_{B,C}(R)$

\verb|select distinct B, C from R|
\subsubsection{Renaming}
$\rho_{\{B\mapsto E,D\mapsto F\}}(R)$

\verb|select A, B as E, c, D as F from R|
\subsubsection{Union}
$R\cup S$

\verb|(select * from R) union (select * from S)|
\subsubsection{Intersection}
$R\cap S$

\verb|(select * from R) intersect (select * from S)|

\subsubsection{Difference}
$R - S$

\verb|(select * from R) except (select * from S)|
\subsubsection{Product}
$R\times S$

\verb|select A, B, C, D from R cross join S|

\verb|select A, B, C, D from R, S|

Note that after crossing, the pair of pairs is flattened
\subsubsection{Natural Join}
\textbf{Notation}

Relational Schema: $R(A)$, where $A=\{A_1, A_2, ...,A_n\}$ are the attribute names.

$R(A,B)$ means $R(A\cup B)$ and $A\cap B=\phi$

$u.[A]=v.[A]$ abbreviates $(u.A_1=v.A_1)\land...\land (u.A_n=v.A_n)$
\\\\
Given $R(A,B)$ and $S(B,C)$, we define natural join as a relation over attributes $A, B, C$ as
\begin{align*}
R\bowtie S\equiv \{t|\exists u\in R,v\in S,u.[B]=v.[B]\land t=u.[A]\cup u.[B] \cup v.[C]\}\\
R\bowtie S=\pi_{A,B,C}(\sigma_{B=B'}(R\times\rho_{B\mapsto B'}(S)))
\end{align*}
So natural join is a composite function.
\section{Implementing ER with SQL}
The entity-relationship model does not dictate implementation. However some methods are better than others...
\\\\
Lumping all data in one big table leads to:
\begin{itemize}
\item Insertion anomalies: 
	\begin{itemize}
	\item if some of the columns of the data to be inserted are supposed to have the same values as some of the columns in the table already, how do we check if they are the same?
	\item Insert data where some columns are empty
	\end{itemize}	 
\item Deletion anomalies:
	\begin{itemize}
	\item deleting a row will delete more information than we should
	\item the other method is to make those columns null
	\end{itemize}
\item Update anomalies: 
	\begin{itemize}
	\item Many rows to update if the rows share some column values
	\end{itemize}
\end{itemize}
\begin{itemize}
\item a transaction might take very long are it needs to scan through multiple rows when information is duplicated in the table (i.e. checks for correctness)
\item this database cannot support many concurrent updates
\end{itemize}
A better idea is to break tables down, having tables for both entities and relationships. Then to retrieve the table, we can join the smaller tables up.
\begin{itemize}
\item Upside: reduce update anomalies
\item Downside: a lot of effort to combine information when making queries
\end{itemize}
\subsection{Keys}
To make sure that the relational implementation is correct, we can use keys and foreign keys.
\\\\
Suppose $R(X)$ is a relational schema with $Z\subseteq X$. If for any records $u$ and $v$ in any instance of $R$ we have
\begin{align}
u.[Z]=v.[Z]\Rightarrow u.[X]=v.[X]
\end{align}
(i.e. Z uniquely identifies a single record), then Z is a superkey for R. If no proper subset of Z is a superkey, (i.e. Z is the smallest combination of attributes needed to uniquely identify a record), then Z is a key for R. We write $R(\underline{Z},Y)$ to indicate that Z is a key for $R(Z\cup Y$).
\\\\
Suppose we have $R(\underline{Z},Y)$. Furthermore, let $S(W)$ be a relational schema with $Z\subseteq W$. We say that Z represents a foreign key in S for R if for any instance we have $\pi_Z(S)\subseteq \pi_Z(R)$ (i.e. all records of attribute Z in S point to a record with the same attribute Z in R).
\\\\
A database is said to have referential integrity when all foreign key constraints are satisfied.
\\\\
\begin{lstlisting}[language=sql]
create table genres (
	genre_id integer NOT NULL,
	genre varchar(100) NOT NULL,
	PRIMARY KEY (genre_id));

create table has_genre (
	movie_id varchar(16) NOT NULL
		REFERENCES movies (movie_id),
	genre_id integer NOT NULL
		REFERENCES genres (genre_id),
	PRIMARY KEY (movie_id, genre_id));
\end{lstlisting}
SQL allows you to specify these constraints, and will prevent any violations of referential integrity (i.e. preventing you from deleting a row with a primary key when another table references that primary key as a foreign key)
\subsection{Implementing Relationships}
\begin{itemize}
\item Many to many: $R(\underline{X,Z},U)$
\item one to many: $R(\underline{X},Z,U)$
\item many to one: $R(X, \underline{Z}, U)$
\end{itemize}
In a many to many relationship, it is intuitive that you can identify each record only you have the details of X and Z.
\\\\
In a one to many relationship (i.e. one department to many employees), then for each employee the department may be repeated so only the employee's identification should be a primary key.
\\\\
Following the previous example, we might want to include the relationship between employees and departments into the table for employees. This is especially so if the entity is a weak entity.
\begin{itemize}
\item Pros: Less complex queries
\item Cons: If the schema needs to change (i.e. one to many to many to many) migration is needed, instead of just needing to add more records to the relationship table.
\end{itemize}
Multiple relationships in a single table
\\\\
If S and T are related by 2 relationships, a workaround could be to have one relationship table, with an indicator variable to specify whether we are referring to the first or second relationship. Then we have 2 more columns, one of them null and the other is the value of the correct relationship.
\begin{align}
\text{Instead of:}\\
R(\underline{X,Z},U)\\
Q(\underline{X,Z},V)\\
\text{We could have:}\\
RQ(\underline{X,Z,type}, U, V)
\end{align}
\section{SQL Commands}
\textbf{Aggregate Commands}
\begin{lstlisting}[language=SQL]
select position, count(*) as total
from has_position
group by position
order by total desc;
\end{lstlisting}
SQL creates new rows where each row 'contains' all the rows which have the same \verb|position|. Then, each column in the new table can only be aggregate commands like \verb|count(*)|, \verb|min(...)|.
\\\\
\textbf{Join}
\begin{lstlisting}[language=SQL]
select title, genre
from movies
join has_genre on has_genre.movie_id = movies.movie_id
join genres on genres.genre_id = has_genre.genre_id
where year = 2017
limit 20;


\end{lstlisting}
\begin{lstlisting}
TITLE                   GENRE
----------------------  ---------
Wonder Woman            Fantasy
Wonder Woman            Action
Wonder Woman            Adventure
It                      Horror
The Greatest Showman    Drama
The Greatest Showman    Musical
The Greatest Showman    Biography
The Foreigner           Action
The Foreigner           Thriller
A Dog's Purpose         Drama
A Dog's Purpose         Comedy
A Dog's Purpose         Adventure
Blade Runner 2049       Drama
Blade Runner 2049       Action
Blade Runner 2049       Mystery
Spider-Man: Homecoming  Action
Spider-Man: Homecoming  Sci-Fi
Spider-Man: Homecoming  Adventure
Coco                    Animation
Coco                    Adventure
\end{lstlisting}
Joining duplicates entries. Since some movies might have more than 1 genre, identifying these movies requires work as SQL evaluates each row separately so \verb|select * from movies join genres where genre = genre_1 and genre = genre_2| doesn't work.
\\\\
\textbf{Multiple Joins}

The way around this problem is to just join the \verb|genres| table multiple times so each row has two \verb|genres| columns!
\begin{lstlisting}[language=sql]
select title, year, rating, votes
from movies as m
join has_genre as hg1 on hg1.movie_id = m.movie_id
join has_genre as hg2 on hg2.movie_id = m.movie_id
join genres as g1 on g1.genre_id = hg1.genre_id
join genres as g2 on g2.genre_id = hg2.genre_id
where m.votes > 100000 and g1.genre = 'Romance' and g2.genre = 'Comedy'
order by votes desc;
\end{lstlisting}
If you are discerning enough, you'll realise that there are actually 4 rows for each original row! This results in a bigger search time complexity. Example rows are shown below.
\begin{lstlisting}
TITLE                 GENRE      GENRE
--------------------  ---------  ---------
In the Mood for Love  Drama      Drama
In the Mood for Love  Drama      Romance
In the Mood for Love  Romance    Drama
In the Mood for Love  Romance    Romance
Chicken Run           Animation  Animation
Chicken Run           Animation  Comedy
Chicken Run           Animation  Adventure
Chicken Run           Comedy     Animation
Chicken Run           Comedy     Comedy
Chicken Run           Comedy     Adventure
\end{lstlisting}
\textbf{Nesting}

Possible to nest queries, i.e. search a sample space generated from another select statement.
\begin{lstlisting}[language=sql]
select m1.title, m1.year, g.genre, m1.rating, m1.votes
from movies as m1
join has_genre as hg on hg.movie_id = m1.movie_id
join genres as g on g.genre_id = hg.genre_id
where m1.votes > 100000 and (not (g.genre = 'Romance' or g.genre = 'Comedy'))
      and m1.movie_id in
         (select m2.movie_id
         from movies as m2
         join has_genre as hg1 on hg1.movie_id = m2.movie_id
         join has_genre as hg2 on hg2.movie_id = m2.movie_id
         join genres as g1 on g1.genre_id = hg1.genre_id
         join genres as g2 on g2.genre_id = hg2.genre_id
         where g1.genre = 'Romance' and g2.genre = 'Comedy')
order by m1.votes desc
limit 10;
\end{lstlisting}
\textbf{Views}

Instead of nesting select statements, it is possible to create a \verb|view|, which can be used in the same way as a table.
\begin{lstlisting}[language=sql]
drop view if exists romcom_ids;

create view romcom_ids as
    select m.movie_id as movie_id
    from movies as m
    join has_genre as hg1 on hg1.movie_id = m.movie_id
    join has_genre as hg2 on hg2.movie_id = m.movie_id
    join genres as g1 on g1.genre_id = hg1.genre_id
    join genres as g2 on g2.genre_id = hg2.genre_id
    where g1.genre = 'Romance' and g2.genre = 'Comedy';

select m.title, m.year, g.genre, m.rating, m.votes
from romcom_ids as r
join has_genre as hg on hg.movie_id = r.movie_id
join genres as g on g.genre_id = hg.genre_id
join movies as m on m.movie_id = r.movie_id
where m.votes > 100000 and (not (g.genre = 'Romance' or g.genre = 'Comedy'))
order by m.votes desc
limit 10;
\end{lstlisting}
\textbf{Three-valued Logic}

SQL allows columns to be \verb|null|. This means that statements can evaluate to \verb|true, false| and \verb|null|.
\begin{lstlisting}[language=sql]
select count(*)
from people
where deathYear = null

returns 0
\end{lstlisting}
\verb|* = null| returns null, and \verb|select ... where| only returns records when the \verb|where| clause returns \verb|true|. 
\\\\
To solve this, SQL introduced \verb|is null|.
\begin{lstlisting}[language=sql]
select count(*)
from people
where deathYear is null

returns 5919
\end{lstlisting}
\textbf{Left Join}

The way \verb|join| works is to strike off the rows which cannot match the other table. However, in some situations we would like to preserve these rows, and we can do so using \verb|left join| or \verb|right join|.
\begin{lstlisting}[language=sql]
select name, position
from people as p
join has_position as c on p.person_id = c.person_id
join movies as m on c.movie_id = m.movie_id
where title = 'Silver Linings Playbook';
\end{lstlisting}
\begin{lstlisting}
NAME               POSITION
-----------------  --------
Robert De Niro     actor
Bruce Cohen        producer
Bradley Cooper     actor
Donna Gigliotti    producer
Jonathan Gordon    producer
David O. Russell   director
Jacki Weaver       actor
Jennifer Lawrence  actor
Matthew Quick      writer
\end{lstlisting}
Joining this table with the roles table gives:
\begin{lstlisting}[language=sql]
select name, position, role
from people as p
join has_position as c on p.person_id = c.person_id
join movies as m on c.movie_id = m.movie_id
join plays_role as r on r.movie_id = m.movie_id and r.person_id = c.person_id
where title = 'Silver Linings Playbook';
\end{lstlisting}
\begin{lstlisting}
NAME               POSITION  ROLE
-----------------  --------  -------
Robert De Niro     actor     Pat Sr.
Bradley Cooper     actor     Pat
Jacki Weaver       actor     Dolores
Jennifer Lawrence  actor     Tiffany
\end{lstlisting}
So some entries in the first table are not shown because only actors have roles. In this case, using a left join will help:
\begin{lstlisting}[language=sql]
select name, position, role
from people as p
join has_position as c on p.person_id = c.person_id
join movies as m on c.movie_id = m.movie_id
left join plays_role as r on r.movie_id = m.movie_id and r.person_id = c.person_id
where title = 'Silver Linings Playbook';
\end{lstlisting}
\begin{lstlisting}
NAME               POSITION  ROLE
-----------------  --------  -------
Robert De Niro     actor     Pat Sr.
Bruce Cohen        producer  [null]
Bradley Cooper     actor     Pat
Donna Gigliotti    producer  [null]
Jonathan Gordon    producer  [null]
David O. Russell   director  [null]
Jacki Weaver       actor     Dolores
Jennifer Lawrence  actor     Tiffany
Matthew Quick      writer    [null]
\end{lstlisting}
Note that \verb|null| is displayed as \verb|[null]| to distinguish between the empty string and actual \verb|null|.
\\\\
\textbf{Select Distinct}

SQL is based on multisets of records (i.e. sets containing duplicates, also called bags) so the following query will result in multiple repeated rows:
\begin{lstlisting}[language=sql]
select r1.role as role, m1.title as title, m1.year as year
from plays_role as r1
join plays_role as r2 on r2.person_id = r1.person_id
join movies as m1 on m1.movie_id = r1.movie_id
join movies as m2 on m2.movie_id = r2.movie_id
join people as p on p.person_id = r1.person_id
where p.name = 'Noomi Rapace'
      and r1.role = r2.role
      and m1.movie_id <> m2.movie_id
order by m1.title, r1.role, m1.year;
\end{lstlisting}
\begin{lstlisting}
ROLE              TITLE                                  YEAR
----------------  -------------------------------------  ----
Lisbeth Salander  The Girl Who Kicked the Hornet's Nest  2009
Lisbeth Salander  The Girl Who Kicked the Hornet's Nest  2009
Lisbeth Salander  The Girl Who Played with Fire          2009
Lisbeth Salander  The Girl Who Played with Fire          2009
Lisbeth Salander  The Girl with the Dragon Tattoo        2009
Lisbeth Salander  The Girl with the Dragon Tattoo        2009
\end{lstlisting}
They keyword \verb|select distinct| collapses the repeated rows. Note that the table with repeated listings is first generated, then the repeats removed, so there is no improvement in query time (I believe?)
\begin{lstlisting}[language=sql]
select distinct r1.role as role, m1.title as title, m1.year as year
from plays_role as r1
join plays_role as r2 on r2.person_id = r1.person_id
join movies as m1 on m1.movie_id = r1.movie_id
join movies as m2 on m2.movie_id = r2.movie_id
join people as p on p.person_id = r1.person_id
where p.name = 'Noomi Rapace'
      and r1.role = r2.role
      and m1.movie_id <> m2.movie_id
order by m1.title, r1.role, m1.year;
\end{lstlisting}
\begin{lstlisting}
ROLE              TITLE                                  YEAR
----------------  -------------------------------------  ----
Lisbeth Salander  The Girl Who Kicked the Hornet's Nest  2009
Lisbeth Salander  The Girl Who Played with Fire          2009
Lisbeth Salander  The Girl with the Dragon Tattoo        2009
\end{lstlisting}
\end{document}